cardinalR: Generating interesting high-dimensional data structures

### 2025-09-20 paper-cardinalR.Rmd
Added equations in new line

The Mobius strip is generated using the `geozoo::mobius` function with `p = 3`, and `n` defined as above. Each point is sampled from a parameterization defined by an angle $\theta \sim U(0, 2\pi)$, which specifies the position along the circular loop, and a width $w \sim U(-0.4, 0.4)$, which offsets the point across the strip’s band. The cartesian coordinates are $X_1 = \left(1 + (w/2) \cos(\theta/2)\right) \cos(\theta)$, $X_2 = \left(1 + (w/2) \cos(\theta/2)\right) \sin(\theta)$, and $X_3 = (w/2)\sin(\theta/2)$. This maps a $2\text{-}D$ band with a half-twist into $3\text{-}D$ space, forming a non-orientable one-sided surface. For $p > 3$, additional noise dimension are added to embed the $3\text{-}D$ Mobius into $p\text{-}D$.

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, `gen_pyrrect(n, p, h, l_vec, rt)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) a"} else { "Figure \\@ref(fig:pyr-proj) a"})`), let $r_x(z)$ and $r_y(z)$ denote the half-widths of the rectangular cross-section at height $z$. That is, $r_x(z) = r_t + (l_x - r_t)z/h$, $r_y(z) = r_t + (l_y - r_t)z/h$. The first three coordinates are then defined as: $X_1 \sim U(-r_x(z),\ r_x(z))$, $X_2 \sim U(-r_y(z),\ r_y(z))$, and $X_3 \sim U(-r_x(z),\ r_x(z)).$

@@

For the S-curve structure, `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended to embed the structure in higher dimensions.

@@

The simplest case, `gen_circle(n, p)` creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) a"} else { "Figure \\@ref(fig:sphere-proj) a"})`). Let a latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2)\pi/2p$. Each additional dimension is computed as: $X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

Building on simple circular structures, the `gen_unifsphere(n, p, r)` function function extends the idea to three dimensions by generating $n$ observations approximately uniformly distributed on the surface of a sphere of radius $r$, with optional high-dimensional noise when $p > 3$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) c"} else { "Figure \\@ref(fig:sphere-proj) c"})`). Each observation is computed from spherical coordinates, with $u \sim U(-1, 1)$ representing $\cos(\phi)$ and $\theta \sim U(0, 2\pi)$ the azimuthal angle. Cartesian coordinates are then defined as $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, and $X_3 = ru$, ensuring uniform distribution on the surface (not within) of the sphere. For $p > 3$, additional dimensions $X_4$ through $X_p$ are appended as structured noise.

@@

Finally, the `gen_hemisphere(n, p)` function restricts sampling to a hemisphere of a $4\text{-}D$ sphere (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) f"} else { "Figure \\@ref(fig:sphere-proj) f"})`). Using spherical coordinates, the azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional noise dimensions are added to embed the structure in higher dimensions.

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, p, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $X_1 = t \cos(t)$, $X_2 = t \sin(t)$, $X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)$. For $p > 3$, the remaining $p - 3$ dimensions are filled with noise to simulate high-dimensional complexity.

@@

First, the `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) a"} else { "Figure \\@ref(fig:triginometric-proj) a"})`). Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$.

@@

For the $4\text{-}D$ trefoil knot, the function `gen_trefoil4d(n, p, steps)` generates the structure on the $3$-sphere ($S^3 \subset \mathbb{R}^4$) using two angular parameters, $\theta$ and $\phi$. A band of thickness around the knot path is controlled by the `steps` argument, while the number of $\theta$ and $\phi$ values is determined by the `steps` and `n` arguments, respectively (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) a"} else { "Figure \\@ref(fig:trefoil-proj) a"})`). The coordinates are defined as $X_1 = \cos(\theta) \cos(\phi)$, $X_2 = \cos(\theta) \sin(\phi)$, $X_3 = \sin(\theta) \cos(1.5 \phi)$, and $X_4 = \sin(\theta) \sin(1.5 \phi)$, where $\theta$ and $\phi$ trace the knot’s path. For $p > 4$, the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the $3\text{-}D$ stereographic projection, `gen_trefoil3d(n, p, steps)` maps each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $(X_1', X_2', X_3') \in \mathbb{R}^3$ using $X_1' = X_1 / (1 - X_4)$, $X_2' = X_2 / (1 - X_4)$, and $X_3' = X_3 / (1 - X_4)$, excluding points where $X_4 = 1$ to avoid division by zero (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) b"} else { "Figure \\@ref(fig:trefoil-proj) b"})`). As with the $4\text{-}D$ case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.

>>>>

The Mobius strip is generated using the `geozoo::mobius` function with `p = 3`, and `n` defined as above. Each point is sampled from a parameterization defined by an angle $\theta \sim U(0, 2\pi)$, which specifies the position along the circular loop, and a width $w \sim U(-0.4, 0.4)$, which offsets the point across the strip’s band. The cartesian coordinates are $$X_1 = \left(1 + (w/2) \cos(\theta/2)\right) \cos(\theta),  \quad X_2 = \left(1 + (w/2) \cos(\theta/2)\right) \sin(\theta),\text{ and }X_3 = (w/2)\sin(\theta/2).$$ This maps a $2\text{-}D$ band with a half-twist into $3\text{-}D$ space, forming a non-orientable one-sided surface. For $p > 3$, additional noise dimension are added to embed the $3\text{-}D$ Mobius into $p\text{-}D$.

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, `gen_pyrrect(n, p, h, l_vec, rt)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) a"} else { "Figure \\@ref(fig:pyr-proj) a"})`), let $r_x(z)$ and $r_y(z)$ denote the half-widths of the rectangular cross-section at height $z$. That is, $r_x(z) = r_t + (l_x - r_t)z/h$, $r_y(z) = r_t + (l_y - r_t)z/h$. The first three coordinates are then defined as: $$X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z)).$$

@@

For the S-curve structure, `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $$X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended to embed the structure in higher dimensions.

@@

The simplest case, `gen_circle(n, p)` creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) a"} else { "Figure \\@ref(fig:sphere-proj) a"})`). Let a latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $$X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2)\pi/2p$. Each additional dimension is computed as: $X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

Building on simple circular structures, the `gen_unifsphere(n, p, r)` function function extends the idea to three dimensions by generating $n$ observations approximately uniformly distributed on the surface of a sphere of radius $r$, with optional high-dimensional noise when $p > 3$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) c"} else { "Figure \\@ref(fig:sphere-proj) c"})`). Each observation is computed from spherical coordinates, with $u \sim U(-1, 1)$ representing $\cos(\phi)$ and $\theta \sim U(0, 2\pi)$ the azimuthal angle. Cartesian coordinates are then defined as $$X_1 = r\sqrt{1 - u^2}\cos(\theta), \quad X_2 = r\sqrt{1 - u^2}\sin(\theta),\text{ and }X_3 = ru,$$ ensuring uniform distribution on the surface (not within) of the sphere. For $p > 3$, additional dimensions $X_4$ through $X_p$ are appended as structured noise.

@@

Finally, the `gen_hemisphere(n, p)` function restricts sampling to a hemisphere of a $4\text{-}D$ sphere (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:sphere) f"} else { "Figure \\@ref(fig:sphere-proj) f"})`). Using spherical coordinates, the azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $$X_1 = \sin(\theta_1)\cos(\theta_2), \quad X_2 = \sin(\theta_1)\sin(\theta_2), \quad X_3 = \cos(\theta_1)\cos(\theta_3), \quad X_4 = \cos(\theta_1)\sin(\theta_3).$$ This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional noise dimensions are added to embed the structure in higher dimensions.

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, p, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $$X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).$$ For $p > 3$, the remaining $p - 3$ dimensions are filled with noise to simulate high-dimensional complexity.

@@

First, the `gen_crescent(n, p)` function generates a $p$-dimensional dataset of $n$ observations based on a $2\text{-}D$ crescent-shaped manifold with optional structured high-dimensional noise (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:triginometric) a"} else { "Figure \\@ref(fig:triginometric-proj) a"})`). Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $$X_1 = \cos(\theta), \quad X_2 = \sin(\theta).$$

@@

For the $4\text{-}D$ trefoil knot, the function `gen_trefoil4d(n, p, steps)` generates the structure on the $3$-sphere ($S^3 \subset \mathbb{R}^4$) using two angular parameters, $\theta$ and $\phi$. A band of thickness around the knot path is controlled by the `steps` argument, while the number of $\theta$ and $\phi$ values is determined by the `steps` and `n` arguments, respectively (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) a"} else { "Figure \\@ref(fig:trefoil-proj) a"})`). The coordinates are defined as $$X_1 = \cos(\theta) \cos(\phi), \quad X_2 = \cos(\theta) \sin(\phi), \quad X_3 = \sin(\theta) \cos(1.5 \phi),\text{ and }X_4 = \sin(\theta) \sin(1.5 \phi)$$, where $\theta$ and $\phi$ trace the knot’s path. For $p > 4$, the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the $3\text{-}D$ stereographic projection, `gen_trefoil3d(n, p, steps)` maps each point $(X_1, X_2, X_3, X_4) \in \mathbb{R}^4$ to $$(X_1', X_2', X_3') \in \mathbb{R}^3\text{ using }X_1' = X_1 / (1 - X_4), \quad X_2' = X_2 / (1 - X_4),\text{ and }X_3' = X_3 / (1 - X_4),$$ excluding points where $X_4 = 1$ to avoid division by zero (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:trefoil) b"} else { "Figure \\@ref(fig:trefoil-proj) b"})`). As with the $4\text{-}D$ case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.

### 2025-09-20 paper-cardinalR.tex
Added equations in new line

The Mobius strip is generated using the \texttt{geozoo::mobius} function with \texttt{p\ =\ 3}, and \texttt{n} defined as above. Each point is sampled from a parameterization defined by an angle \(\theta \sim U(0, 2\pi)\), which specifies the position along the circular loop, and a width \(w \sim U(-0.4, 0.4)\), which offsets the point across the strip's band. The cartesian coordinates are \(X_1 = \left(1 + (w/2) \cos(\theta/2)\right) \cos(\theta)\), \(X_2 = \left(1 + (w/2) \cos(\theta/2)\right) \sin(\theta)\), and \(X_3 = (w/2)\sin(\theta/2)\). This maps a \(2\text{-}D\) band with a half-twist into \(3\text{-}D\) space, forming a non-orientable one-sided surface. For \(p > 3\), additional noise dimension are added to embed the \(3\text{-}D\) Mobius into \(p\text{-}D\).

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} (Figure \ref{fig:pyr-proj} a), let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\). That is, \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as: \(X_1 \sim U(-r_x(z),\ r_x(z))\), \(X_2 \sim U(-r_y(z),\ r_y(z))\), and \(X_3 \sim U(-r_x(z),\ r_x(z)).\)

@@

For the S-curve structure, \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended to embed the structure in higher dimensions.

@@

The simplest case, \texttt{gen\_circle(n,\ p)} creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (Figure \ref{fig:sphere-proj} a). Let a latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \(X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\). For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2)\pi/2p\). Each additional dimension is computed as: \(X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

Building on simple circular structures, the \texttt{gen\_unifsphere(n,\ p,\ r)} function function extends the idea to three dimensions by generating \(n\) observations approximately uniformly distributed on the surface of a sphere of radius \(r\), with optional high-dimensional noise when \(p > 3\) (Figure \ref{fig:sphere-proj} c). Each observation is computed from spherical coordinates, with \(u \sim U(-1, 1)\) representing \(\cos(\phi)\) and \(\theta \sim U(0, 2\pi)\) the azimuthal angle. Cartesian coordinates are then defined as \(X_1 = r\sqrt{1 - u^2}\cos(\theta)\), \(X_2 = r\sqrt{1 - u^2}\sin(\theta)\), and \(X_3 = ru\), ensuring uniform distribution on the surface (not within) of the sphere. For \(p > 3\), additional dimensions \(X_4\) through \(X_p\) are appended as structured noise.

@@

Finally, the \texttt{gen\_hemisphere(n,\ p)} function restricts sampling to a hemisphere of a \(4\text{-}D\) sphere (Figure \ref{fig:sphere-proj} f). Using spherical coordinates, the azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \(X_1 = \sin(\theta_1)\cos(\theta_2)\), \(X_2 = \sin(\theta_1)\sin(\theta_2)\), \(X_3 = \cos(\theta_1)\cos(\theta_3)\), \(X_4 = \cos(\theta_1)\sin(\theta_3)\). This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional noise dimensions are added to embed the structure in higher dimensions.

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ p,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \(X_1 = t \cos(t)\), \(X_2 = t \sin(t)\), \(X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)\). For \(p > 3\), the remaining \(p - 3\) dimensions are filled with noise to simulate high-dimensional complexity.

@@

First, the \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise (Figure \ref{fig:triginometric-proj} a). Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \(X_1 = \cos(\theta)\), \(X_2 = \sin(\theta)\).

@@

For the \(4\text{-}D\) trefoil knot, the function \texttt{gen\_trefoil4d(n,\ p,\ steps)} generates the structure on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)) using two angular parameters, \(\theta\) and \(\phi\). A band of thickness around the knot path is controlled by the \texttt{steps} argument, while the number of \(\theta\) and \(\phi\) values is determined by the \texttt{steps} and \texttt{n} arguments, respectively (Figure \ref{fig:trefoil-proj} a). The coordinates are defined as \(X_1 = \cos(\theta) \cos(\phi)\), \(X_2 = \cos(\theta) \sin(\phi)\), \(X_3 = \sin(\theta) \cos(1.5 \phi)\), and \(X_4 = \sin(\theta) \sin(1.5 \phi)\), where \(\theta\) and \(\phi\) trace the knot's path. For \(p > 4\), the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the \(3\text{-}D\) stereographic projection, \texttt{gen\_trefoil3d(n,\ p,\ steps)} maps each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), excluding points where \(X_4 = 1\) to avoid division by zero (Figure \ref{fig:trefoil-proj} b). As with the \(4\text{-}D\) case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.

>>>>

The Mobius strip is generated using the \texttt{geozoo::mobius} function with \texttt{p\ =\ 3}, and \texttt{n} defined as above. Each point is sampled from a parameterization defined by an angle \(\theta \sim U(0, 2\pi)\), which specifies the position along the circular loop, and a width \(w \sim U(-0.4, 0.4)\), which offsets the point across the strip's band. The cartesian coordinates are \[X_1 = \left(1 + (w/2) \cos(\theta/2)\right) \cos(\theta),  \quad X_2 = \left(1 + (w/2) \cos(\theta/2)\right) \sin(\theta),\text{ and }X_3 = (w/2)\sin(\theta/2).\] This maps a \(2\text{-}D\) band with a half-twist into \(3\text{-}D\) space, forming a non-orientable one-sided surface. For \(p > 3\), additional noise dimension are added to embed the \(3\text{-}D\) Mobius into \(p\text{-}D\).

@@

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} (Figure \ref{fig:pyr-proj} a), let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\). That is, \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as: \[X_1 \sim U(-r_x(z),\ r_x(z)), \quad X_2 \sim U(-r_y(z),\ r_y(z)),\text{ and }X_3 \sim U(-r_x(z),\ r_x(z)).\]

@@

For the S-curve structure, \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \[X_1 = \sin(\theta), \quad X_2 \sim U(0, 2), \quad X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\]. This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended to embed the structure in higher dimensions.

@@

The simplest case, \texttt{gen\_circle(n,\ p)} creates a unit circle in two dimensions, with the remaining dimensions forming sinusoidal extensions of the angular parameter at progressively smaller scales (Figure \ref{fig:sphere-proj} a). Let a latent angle variable \(\theta\) is uniformly sampled from the interval \([0, 2\pi]\). Coordinates in the first two dimensions represent a perfect circle on the plane: \[X_1 = \cos(\theta), \quad X_2 = \sin(\theta)\]. For dimensions \(X_3\) through \(X_p\), sinusoidal transformations of the angle \(\theta\) are introduced. The first component is a scaling factor that decreases with the dimension index, defined as \(\text{scale}_j = \sqrt{(0.5)^{j-2}}\) for \(j = 3, \dots, p\). The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula \(\phi_j = (j - 2)\pi/2p\). Each additional dimension is computed as: \(X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p\).

@@

Building on simple circular structures, the \texttt{gen\_unifsphere(n,\ p,\ r)} function function extends the idea to three dimensions by generating \(n\) observations approximately uniformly distributed on the surface of a sphere of radius \(r\), with optional high-dimensional noise when \(p > 3\) (Figure \ref{fig:sphere-proj} c). Each observation is computed from spherical coordinates, with \(u \sim U(-1, 1)\) representing \(\cos(\phi)\) and \(\theta \sim U(0, 2\pi)\) the azimuthal angle. Cartesian coordinates are then defined as \[X_1 = r\sqrt{1 - u^2}\cos(\theta), \quad X_2 = r\sqrt{1 - u^2}\sin(\theta),\text{ and }X_3 = ru,\] ensuring uniform distribution on the surface (not within) of the sphere. For \(p > 3\), additional dimensions \(X_4\) through \(X_p\) are appended as structured noise.

@@

Finally, the \texttt{gen\_hemisphere(n,\ p)} function restricts sampling to a hemisphere of a \(4\text{-}D\) sphere (Figure \ref{fig:sphere-proj} f). Using spherical coordinates, the azimuthal angle \(\theta_1 \sim U(0, \pi)\) in the \((x_1, x_2)\) plane, while the elevation angle \(\theta_2 \sim U(0, \pi)\) in the \((x_2, x_3)\) plane. Additionally, \(\theta_3 \sim U(0, \pi/2)\) in the \((x_3, x_4)\) plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into \(4\text{-}D\) Cartesian space: \[X_1 = \sin(\theta_1)\cos(\theta_2), \quad X_2 = \sin(\theta_1)\sin(\theta_2), \quad X_3 = \cos(\theta_1)\cos(\theta_3), \quad X_4 = \cos(\theta_1)\sin(\theta_3).\] This produces points on one side of a \(4\text{-}D\) unit sphere, effectively generating a \(4\text{-}D\) hemisphere. For \(p > 4\), additional noise dimensions are added to embed the structure in higher dimensions.

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ p,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \[X_1 = t \cos(t), \quad X_2 = t \sin(t), \quad X_3 \sim U(w_1, w_2), \text{ where } t \sim U(0, 3\pi).\] For \(p > 3\), the remaining \(p - 3\) dimensions are filled with noise to simulate high-dimensional complexity.

@@

First, the \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise (Figure \ref{fig:triginometric-proj} a). Let \(\theta \in [\pi/6, 2\pi]\) be a sequence of \(n\) evenly spaced angles. The corresponding \(2\text{-}D\) coordinates are defined by: \[X_1 = \cos(\theta), \quad X_2 = \sin(\theta).\]

@@

For the \(4\text{-}D\) trefoil knot, the function \texttt{gen\_trefoil4d(n,\ p,\ steps)} generates the structure on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)) using two angular parameters, \(\theta\) and \(\phi\). A band of thickness around the knot path is controlled by the \texttt{steps} argument, while the number of \(\theta\) and \(\phi\) values is determined by the \texttt{steps} and \texttt{n} arguments, respectively (Figure \ref{fig:trefoil-proj} a). The coordinates are defined as \[X_1 = \cos(\theta) \cos(\phi), \quad X_2 = \cos(\theta) \sin(\phi), \quad X_3 = \sin(\theta) \cos(1.5 \phi),\text{ and }X_4 = \sin(\theta) \sin(1.5 \phi)\], where \(\theta\) and \(\phi\) trace the knot's path. For \(p > 4\), the quadratic structure is embedded into higher dimensions by appending additional noise dimensions.

@@

For the \(3\text{-}D\) stereographic projection, \texttt{gen\_trefoil3d(n,\ p,\ steps)} maps each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \[(X_1', X_2', X_3') \in \mathbb{R}^3\text{ using }X_1' = X_1 / (1 - X_4), \quad X_2' = X_2 / (1 - X_4),\text{ and }X_3' = X_3 / (1 - X_4),\] excluding points where \(X_4 = 1\) to avoid division by zero (Figure \ref{fig:trefoil-proj} b). As with the \(4\text{-}D\) case, optional noise dimensions can be added to embed the knot into higher-dimensional spaces.
